---
title: "GAMs Updated March 2026 - Females"
author: "SYC"
date: "2026-03-15"
output: html_document
---

## Packages
```{r}

library(mgcv)
library(gratia)
library(mgcViz)
library(tidygam)


```

## Data Prep
### Load data
#### All FC
```{r}
#| echo: false

aa <- read.csv('/data/all_yeo7_fc.csv')

aa.hc <- read.csv('/data/all_yeo7xHc_fc.csv')

aa.amyg <- read.csv('/data/all_yeo7xAmyg_fc.csv')

aa.thal <- read.csv('/data/all_yeo7xThal_fc.csv')

aa.nac <- read.csv('/data/all_yeo7xNAc_fc.csv')

aa.cau <- read.csv('/data/all_yeo7xCau_fc.csv')

aa.put <- read.csv('/data/all_yeo7xPut_fc.csv')

common_colnames <- intersect(colnames(aa), colnames(aa.hc))
aa.df <- merge(aa, aa.hc, by = c(common_colnames))
aa.df <- merge(aa.df, aa.amyg, by = c(common_colnames))
aa.df <- merge(aa.df, aa.thal, by = c(common_colnames))
aa.df <- merge(aa.df, aa.nac, by = c(common_colnames))
aa.df <- merge(aa.df, aa.cau, by = c(common_colnames))
aa.df <- merge(aa.df, aa.put, by = c(common_colnames))

rm(aa, aa.hc, aa.amyg, aa.thal, aa.nac, aa.cau, aa.put)

```

### Check Ethnicity - No others in testing
```{r}

## Assign others

aa.df[577,3] <- "indian"
aa.df[578,3] <- "indian"


```


#### Demo
```{r}

demo <- read.csv('/data/all_demo.csv')

tr.id <- read.csv('/data/tr.id.csv')
  
te.id <- read.csv('/data/te.id.csv')  


```

### Training
```{r}

aa.tr_nh <- merge(tr.id, aa.df, by = c("subjid", "ds", "dx"), all.x = TRUE)

aa.tr_nh <- aa.tr_nh %>%
  arrange(order)

aa.tr_nh$subjid <- as.factor(aa.tr_nh$subjid)
aa.tr_nh$sex <- as.factor(aa.tr_nh$sex)
aa.tr_nh$ethnicity <- as.factor(aa.tr_nh$ethnicity)
aa.tr_nh$batch <- as.factor(aa.tr_nh$batch)
contrasts(aa.tr_nh$batch) <- contr.sum(length(levels(aa.tr_nh$batch)))

```

#### Testing

```{r}

aa.te_nh <- merge(te.id, aa.df, by = c("subjid", "ds", "dx"), all.x = TRUE)

aa.te_nh <- aa.te_nh %>%
  arrange(order) %>%
  mutate(cohort = ifelse(
     str_detect(ds, "gusto") & (is.na(dx) | dx==0), "gusto_con", ifelse(   
    str_detect(ds, "gusto") & dx == 1, "gusto_case", ifelse(
    ds == "braco", "braco", "spresto"))))

check <- aa.te_nh %>%
  filter(is.na(cohort))

aa.te_nh$subjid <- as.factor(aa.te_nh$subjid)
aa.te_nh$sex <- as.factor(aa.te_nh$sex)
aa.te_nh$ethnicity <- as.factor(aa.te_nh$ethnicity)
aa.te_nh$batch <- as.factor(aa.te_nh$batch)


```


### Save data
```{r}
setwd('/data')

write.csv(aa.tr_nh, "train_dat_all.csv", row.names = FALSE)
write.csv(aa.te_nh, "test_dat_all.csv", row.names = FALSE)

```

## START HERE IF DATA IS PREPPED
```{r}

### note after saving as csv, factors become numeric/characters

aa.tr_nh <- read.csv('/data/train_dat_all.csv')

aa.te_nh <- read.csv('/data/test_dat_all.csv')

test <- aa.te_nh %>%
  filter(ds != "spresto")

```

### Training & Testing
```{r}

aa.tr_nh <- aa.tr_nh %>%
  arrange(order)

aa.tr_nh$subjid <- as.factor(aa.tr_nh$subjid)
aa.tr_nh$sex <- as.factor(aa.tr_nh$sex)
aa.tr_nh$ethnicity <- as.factor(aa.tr_nh$ethnicity)
aa.tr_nh$batch <- as.factor(aa.tr_nh$batch)
contrasts(aa.tr_nh$batch) <- contr.sum(length(levels(aa.tr_nh$batch)))

check <- aa.te_nh %>%
  filter(is.na(cohort))

aa.te_nh$subjid <- as.factor(aa.te_nh$subjid)
aa.te_nh$sex <- as.factor(aa.te_nh$sex)
aa.te_nh$ethnicity <- as.factor(aa.te_nh$ethnicity)
aa.te_nh$batch <- as.factor(aa.te_nh$batch)

### check
#### batch, subjid, ethnicity, sex must be factor
#### ethnicity should only have 3 levels in testing
# str(aa.tr_nh) 
# str(aa.te_nh) 
```

## 1. GAM
### Prep
```{r}

basis.check <- function(b, k.sample = 5000, k.rep = 200) {
  mgcv:::k.check(b, subsample = k.sample, n.rep = k.rep)
}

sample.dat = data.frame(ScanAge = c(4.0, 4.4, 4.8, 5.2, 5.6, 6.0, 6.4, 6.8, 7.2, 7.6, 8.0, 8.4, 8.8, 9.2, 9.6, 10.0, 10.4, 10.8, 11.2, 11.6))

sample.dat$QC_meanmotion <- 0.13
sample.dat$ethnicity <- "chinese"
sample.dat$batch <- 1

sample.dat$ethnicity <- as.factor(sample.dat$ethnicity)
sample.dat$batch <- as.factor(sample.dat$batch)

coln <- colnames(aa.tr_nh)[12:116]
varname <- coln

varname <- str_replace(varname, fixed('1'), "VN")
varname <- str_replace(varname, fixed('2'), "SMN")
varname <- str_replace(varname, fixed('3'), "DAN")
varname <- str_replace(varname, fixed('4'), "VAN")
varname <- str_replace(varname, fixed('5'), "LN")
varname <- str_replace(varname, fixed('6'), "FPN")
varname <- str_replace(varname, fixed('7'), "DMN")

varname <- str_replace(varname, "^y", "")

key <- data.frame(sex = rep("Female", 105), coln = coln, varname = varname)


```

### New Function 
**v March 2026**
++ change method from gaussian to gaulss to model both location and scale (variance)
++ change k to fix k of 6
++ remove plot
++ remove sample.dat prediction for plotting data 
```{r}

gam_gaulss_zscores <- function(sex, coln, varname) {
  data <- aa.tr_nh
  data <- data %>%
    filter(sex == !!sex) %>%
    rename(y = !!coln) 
  test <- aa.te_nh %>%
    filter(sex == !!sex) %>%
    rename(y = !!coln)  
  
  gg <- gam(list(
    y ~ s(ScanAge, k = 6, bs= 'tp') + ethnicity + QC_meanmotion + batch + s(subjid, bs = 'fs'),
    ~ batch + s(ScanAge, k = 6, bs= 'tp') + QC_meanmotion
    ),
    family=gaulss(), data=data, method = "REML")
  
  gg_cprep <- getViz(gg)
  gg_c <- check(gg_cprep,
      a.qq = list(method = "tnorm", 
                  a.cipoly = list(fill = "light blue")), 
      a.respoi = list(size = 0.5), 
      a.hist = list(bins = 10))

  gg_basis <- basis.check(gg)
  
  pred_gg <- predict.gam(gg, newdata = test, type = "link", exclude = "s(subjid)")
  
  predz <- test %>%
    dplyr::select(subjid, sex, ScanAge, y, QC_meanmotion, ethnicity, batch, ds, cohort) %>%
    na.omit() %>%
    mutate(fit = pred_gg[,1]) %>%
    mutate(se = predict.gam(gg, newdata = test, se.fit = TRUE, exclude = "s(subjid)")$se.fit[,1]) %>%
    mutate(sigma = exp(pred_gg[,2])) %>%
    mutate(z = (y - fit)/sigma)  
  
  all_coefs <- coef(gg)
  site_par <- data.frame(
    site = c("batch1", "batch2", "batch3"),
    mean = c(all_coefs["batch1"], all_coefs["batch2"], (-(all_coefs["batch1"] + all_coefs["batch2"]))), 
    scale = c(all_coefs["batch1.1"], all_coefs["batch2.1"], (-(all_coefs["batch1.1"] + all_coefs["batch2.1"])))
  )
  
  key = c(coln, varname)
  
    list("gam" = gg, "z" = predz, "check.plots" = gg_c, "check.basis" = gg_basis, "site_par" = site_par, "key" = key)
    
}
  
```

### Test on one
```{r}

test.basis <- gam_gaulss_zscores("Female", "y1x2", "VNxSMN")

check <- test.basis$z
summary(test.basis$gam)

test.basis$check.plots
test.basis$check.basis
test.basis$site_par

```

### Run on Multiple IDs
++ add forloop to run GAMs over multiple features vs 1 by 1 
```{r}

## can run from 1:105, but ran in steps here to check how long it takes/save output in case of errors/crashes

outlist <- list()

for (i in 1:21) {
  sex_par <- key[i,1]
  coln_par <- key[i,2]
  varname_par <- key[i,3]
  outlist[[varname_par]] <- gam_gaulss_zscores(sex_par, coln_par, varname_par)
}

for (i in 22:35) {
  sex_par <- key[i,1]
  coln_par <- key[i,2]
  varname_par <- key[i,3]
  outlist[[varname_par]] <- gam_gaulss_zscores(sex_par, coln_par, varname_par)
}

for (i in 36:77) {
  sex_par <- key[i,1]
  coln_par <- key[i,2]
  varname_par <- key[i,3]
  outlist[[varname_par]] <- gam_gaulss_zscores(sex_par, coln_par, varname_par)
}

for (i in 78:105) {
  sex_par <- key[i,1]
  coln_par <- key[i,2]
  varname_par <- key[i,3]
  outlist[[varname_par]] <- gam_gaulss_zscores(sex_par, coln_par, varname_par)
}

check2 <- outlist[["VNxSMN"]]
sum.check <- summary(check2$gam)
check2 <- check2$z

```


## 2. Extract Z scores
```{r}

z_f <- data.frame("subjid" = outlist[['VNxSMN']]$z$subjid, "ds" = outlist[['VNxSMN']]$z$ds)

for (i in 1:length(outlist)) {
  temp_name <- names(outlist)[[i]]
  temp_df <- outlist[[temp_name]]$z
  temp_df <- temp_df %>%
    dplyr::select(subjid, ds, z)
  colnames(temp_df)[3] <- paste(temp_name, "z", sep = "_")
  z_f <- merge(z_f, temp_df, by = c("subjid", "ds"))
}


z_f <- z_f %>%
  mutate(sex = "Female")

```

### Save Z 
```{r}

setwd('/output')

write.csv(z_f, "z_testing_females.csv", row.names = FALSE)

```


### Save RDS
```{r}

setwd('/output')

saveRDS(outlist, 
        file = "gaulss_females.rds", 
        compress = "xz")

```
